---
format: html
---

## Phylogenetic Inference with `IQTREE`

The final lab of the semester will provide a brief introduction to estimating phylogenies with SNP data. We will infer relationships among five *Microcebus* spp. using data from Poelstra et al. 2021. Unlike Poelstra et al., however, we will avoid species tree inference in favor of a concatenation-based method of phylogenetic inference implemented in `IQTREE`. This approach takes SNP data from a .vcf file, converts genotypes (e.g., `0/0`, `0/1`, `1/1`) into sequence characters (using standard bases for homozygotes and ambiguity codes like `R` for heterozygotes), squishes all sites into a single alignment, and then uses a model of sequence evolution—along with an efficient pruning algorithm to evaluate likelihood—to search for the tree that maximizes the probability of observing these data. (While a diversity of pruning algorithms have been developed, they typically work by taking an initial tree and iteratively modifying its topology and branch lengths, only keeping changes that increase likelihood until no further improvement is found.) The documentation for `IQTREE` [is unusually clear](https://iqtree.github.io/), and worth your time. 

We begin by creating a new `mamba` environment for `IQTREE` and its dependencies:

```bash
name: iqtree
channels:
  - conda-forge
  - bioconda
dependencies:
  - python=3.10
  - iqtree
  - bcftools
  - tabix
  - git
```
```bash
mamba env create -f iqtree.yaml
```

We will also need a handy utility called `vcf2phylip` for converting `.vcf` files into a format called `PHYLIP`, developed for [a program of the same name] by Joe Felsenstein (my graduate school phylogenetics professor). These plain-text files begin with a line indicating the number of samples (here, 50) and sites (35529) in the sequence alignment. Each row begins with the name of a sample, is followed by a tab space, and then concludes with all sequences in a single row, with missing data represented by the character `N`: 

```bash
50 35529
mleh019_r01_p1h12   GRGCCYCGAGAGC.....
mleh020_r01_p2c06   GGGCCTCGAGAGC.....
```

We can download `vcf2phylip` directly from GitHub using the command `git clone` (note that this will create a subdirectory of the same name wherever it is executed): 

```bash
git clone https://github.com/edgardomortiz/vcf2phylip.git
```

We now need an `.sbatch` script that performs multiple tasks: 1) load the `mamba` environment with `IQTREE` and `bcftools`; 2) use `bcftools` to filter the `.vcf` down to SNPs only and index it; 3) convert this filtered `.vcf` to a `.phy` file with `vcf2phylip.py`; and 4) run `IQTREE` on this output: 

```bash
#SBATCH --account=priority-bioe-591-genomics      
#SBATCH --job-name=iqtree                            
#SBATCH --partition=priority          
#SBATCH --nodes=1                      
#SBATCH --ntasks-per-node=1            
#SBATCH --cpus-per-task=1              
#SBATCH --time=0-01:00:00              
#SBATCH --output=iqtree-%j.out
#SBATCH --error=iqtree-%j.err

module load Mamba/23.11.0-0 # load mamba module
eval "$(conda shell.bash hook)" # make conda / mamba executable
conda activate iqtree # load iqtree environment 
bcftools view -v snps lemurs.vcf.gz -Oz -o lemurs.snps.vcf.gz # filter vcf to SNPs
bcftools index lemurs.snps.vcf.gz # index the vcf
python vcf2phylip/vcf2phylip.py -i lemurs.snps.vcf.gz # convert the vcf to phylip
iqtree -s lemurs.snps.min4.phy -m GTR -bb 500 -nt AUTO # run iqtree
```

In the script above, the first three lines of commands should be annoyingly familiar. The next two lines (`bcftools view` and `bcftools index`) are variants of commands you used earlier this semester. Running `vcf2philip` (on line 6) requires only specifying a path to a `.vcf` file with the `-i` argument; note in this example below that I am submitting the job from outside the `vcf2phylip` subdirectory. The final line (beginning with `iqtree`) takes the input file from `vcf2phylip` with the `-s` flag; by default, the program only includes sites with at least 4 samples, encoding this in the `min4` portion of the file name. The remainder of the command specifies the [generalised time reversible model](https://en.wikipedia.org/wiki/Substitution_model) of sequence evolution with the `-m` flag, indicates you'd like 500 bootstrap replicates with the `-bb` flag, and has the program determine the number of threads to use (`-nt`). (The generalised time reversible model, or GTR, allows each type of nucleotide substitution to have its own rate, along with unequal equilibrium base frequencies; it is flexible and widely used in phylogenetics.)     

On successful execution---as usual, you'll need to tweak your paths---the script above should run for somewhere between 1 and 5 minutes. Assuming no errors, the `.out` file is informative and worth scanning; it describes the tree space search process and the contents of output files. 

To visualize our results, we will want to switch to Tempest Web, opening a new interactive RStudio session with the latest version of `R`. We'll then install `ape`, [a bread-and-butter package for scriptable phylogenetics](https://emmanuelparadis.github.io/): 


```{r, message=FALSE, warning=FALSE}
# install.packages("ape")
library(ape)
library(readr)
```

We use the `read.tree()` function to load the consensus phylogeny output by `IQTREE`: 

```{r, message=FALSE, warning=FALSE}
# read tree
tree <- read.tree("data/lemurs.snps.min4.phy.contree")
tree
```
Attributes of the `tree` object can be accessed with the `$` operator, e.g.:

```{r, message=FALSE, warning=FALSE}
tree$tip.label
```
To make our lives easier, we will read in the the tab-delimited file `lemur_metadata.txt` in the `data/` subdirectory, and overwrite sample ID with its associated species: 

```{r}
# read metadata
meta <- read_tsv("data/lemur_metadata.txt")

# map species to sample
map <- setNames(meta$species, meta$ID_long)

# overwrite tip labels
tree$tip.label <- unname(map[tree$tip.label])
```

The `plot()` function will produce an *unrooted* phylogeny, i.e., one where relationships are not visualized with reference to the samples that are most distantly related to the clades under investigation: 

```{r, message=FALSE, warning=FALSE}
plot(tree)
```

To make this output more interpretable, we use the `root()` function and the  name of *Microcebus murinus* sample. We can also add bootstrapped support values, filtering to show only those that exceed 70 (a common cutoff for confidence in the relationship portrayed by a node): 

```{r, message=FALSE, warning=FALSE}
# root using an outgroup (replace with your sample name)
tree_rooted <- root(tree, outgroup = "murinus", resolve.root = TRUE)

# plot
plot(tree_rooted)

# add node labels
bs <- as.numeric(tree$node.label)
nodelabels(ifelse(bs >= 70, bs, ""), cex = 0.7, frame = "n")
```

This can be exported as a .png with base `R` functions:

```{r, message=FALSE, warning=FALSE}
png("images/lemur_tree.png", width = 1000, height = 800)
plot(tree_rooted)
nodelabels(tree$node.label, cex = 0.7, frame = "n")
dev.off()
```

:::{.callout-note title="**Homework 13**"}

Complete the tutorial above, recording all commands in `bash` or `.sbatch` scripts and an `.R` or `.Rmd` file. In your `.R` script or a separate `.md` file, write 2-3 sentences comparing the phylogeny you inferred above with the results of Poelstra et al. 2021. Does your analysis also support lumping *M. mittermeieri* and *M. lehilahytsara*? Is *Microcebus* sp. 3 still recovered as distinct? Add, commit, and push your work to GitHub. Make your final tweaks to directory structure and your `README.md`. You're officially done with the course!

:::


